---
title: "MATH 201: Lecture 8a Handout"
subtitle: "6.1 Inference for a single proportion"
format: 
  pdf:
    echo: false
    message: false
    warning: false
    geometry:
      - top=0.5in
      - left=1in
      - right=1in
---

```{r}
library(tidyverse)
options(scipen = 999)
```

\newcommand{\blank}{\rule{2.5cm}{0.15mm}}

\vspace*{-2cm}

Name:\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ Date:\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

## Learning goals for today

By the end of this lecture, you should be able to:

- Describe how confidence intervals and hypothesis tests are centered on the idea of a sampling distribution.
- Explain differences between hypothesis tests and confidence intervals.
- Identify whether a confidence interval or a hypothesis test would be most appropriate to answer a given research question.

# Back to Orange Candies

Recall the Reese's Pieces simulation:

- True proportion of orange candies: $p = 0.50$
- We collected a bunch of samples
- For each sample we computed $\hat{p}$ and plotted it

When we repeatedly sampled:

- Did we always get the same $\hat{p}$? 

\vspace{0.5in}

- Where were most $\hat{p}$’s located?  

\vspace{0.5in}

This creates a **sampling distribution of $\hat{p}$**. We now know not to rely on a single sample statistic too heavily, because it could be an unusual value for the sample distribution.

When independence and success failure ($n \times p \geq 10$ and $n \times (1 - p) \geq 10$) conditions are met:

$$
\hat{p} \sim Normal(\mu = \_\_\_\_\_\_\_\_, \sigma = \_\_\_\_\_\_\_\_).
$$

\newpage

# How can we use a single sample?

We take a random sample of $n = 100$ candies and find 56 of them are orange.

## Confidence Interval for a Population Proportion $p$

If we want to estimate $p$, the population proportion, using our sample, we can compute a confidence interval.

Independence condition: We have a random sample, so this is condition should be met.  
Success-failure condition: $n \times \hat{p} = 100 * \frac{56}{100} = 56 \geq 10 \checkmark$ and $n \times (1 - \hat{p}) = 100 * (1 - \frac{56}{100}) = 44 \geq 10 \checkmark$, so this condition is met.

Since conditions are met:
$$\hat{p} \overset{Approximately}{\sim} Normal\left(\mu = p, \sigma = \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}\right).$$

Our 95\% confidence interval:
\begin{equation*}
  \begin{split}
    &\hat{p} \pm 1.96 \times \sqrt{\frac{\hat{p} \times  (1 - \hat{p})}{n}}\\
    &\frac{56}{100} \pm 1.96 \times \sqrt{\frac{\frac{56}{100} \times  (1 - \frac{56}{100})}{100}}\\
    &0.56 \pm 1.96 \times \sqrt{\frac{0.56 \times  (1 - 0.56)}{100}}\\
    &0.56 \pm 1.96 \times 0.049639\\
    &0.56 \pm 0.1\\
    &(0.56 - 0.1, 0.56 + 0.1)\\
    \Rightarrow p \text{ is between } &(0.46, 0.66) \text{ with 95\% confidence.}
  \end{split}
\end{equation*}

Label the curves for $Z\sim N(\mu=0, \sigma = 1)$ and for $\hat{p}\sim N(\mu = p, \sigma = \sqrt{\frac{\hat{p} (1 - \hat{p})}{n}})$, marking: the center, the confidence interval, and the area between the interval upper and lower bounds.

```{r}
#| fig-align: center
#| fig-height: 1
#| fig-width: 4
# Create x values
x <- seq(-4, 4, length.out = 500)

# Standard normal density
df <- data.frame(
  x = x,
  y = dnorm(x)
)

blank_plot <- ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.3) +
  # Only one tick at 0 labeled mu
  # scale_x_continuous(
  #   breaks = 0,
  #   labels = "",
  #   expand = c(0.02, 0.02)
  # ) +
  
  # Remove y-axis ticks and labels
  scale_y_continuous(breaks = NULL) +
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  )

gridExtra::grid.arrange(blank_plot, blank_plot, nrow = 1)
```

\newpage 

## Hypothesis Test for a Population Proportion $p$

Suppose we wanted to test the claim that the true proportion of orange candies, $p$, is not 0.5. We could use a hypothesis test to formally test this with significance level 0.05.

#### Step 1: State Hypotheses

$$
H_0: p = 0.50 \quad \quad H_A: p \neq 0.50
$$

#### Step 2: Check Conditions

Independence condition: We have a random sample, so this is condition should be met.  
Success-failure condition: $n \times p_0 = 100 * 0.5 = 50 \geq 10 \checkmark$ and $n \times (1 - p_0) = 100 * (1 - 0.5) = 50 \geq 10 \checkmark$, so this condition is met.

Since conditions are met:
$$\hat{p} \overset{Approximately}{\sim} Normal\left(\mu = p_0, \sigma = \sqrt{\frac{p_0(1 - p_0)}{n}}\right).$$

#### Step 3: Calculate Evidence

Then we can compute a Z-score:
$$
Z = \frac{\hat{p} - p_0}{\sqrt{\frac{p_0 \times (1 - p_0)}{n}}} = \frac{0.56 - 0.5}{\sqrt{\frac{0.5 \times (1 - 0.5)}{100}}} = \frac{0.06}{0.05} = 1.2
$$

So our sample proportion $\hat{p} = 0.56$ is 1.2 standard deviations above the hypothesized average $p_0 = 0.5$.

We can compute how unlikely results as or more extreme as what we observed are, if the null hypothesis (i.e. $p = 0.5$) were true: 

2 $\times$ (1 - `pnorm(0.56, mean = 0.5, sd = 0.05`) = 2 $\times$ (1 - 0.8849) = 2 $\times$ 0.1151  = 0.2302.

```{r}
#| fig-align: center
#| fig-height: 1
#| fig-width: 2.5
blank_plot
```


```{r}
#| echo: false
#| eval: false
2 * (1 - pnorm(0.56, mean = 0.5, sd = 0.05))
```

#### Step 4: Decide & Conclude

Statistical decision: 

It is not that unlikely to get data like we got, or more extreme, if $H_0$ is true. p-value = 0.2302 > $\alpha = 0.05$ so fail to reject $H_0$.

Conclusion in context:

We failed to find evidence that the true proportion of candies that are orange is not 0.5 (p-value = 0.2302).

\newpage

# Connecting ideas

## Same model, different questions

| Goal | What we do | Key question | Results |
|------|------------|--------------|----------|
| Estimate $p$ | Confidence interval | "What values of $p$ are plausible?" | (0.46, 0.66) |
| Test a claim | Hypothesis test | "Is this value plausible?" | Fail to reject $H_0: p = 0.5$


## Compare the results

Is $0.5$ inside the confidence interval (aka is it a plausible value)?

These are not separate methods—they come from the **same underlying model!**


## What is the SAME? What is DIFFERENT?

Both methods rely on:

- Conditions: independence and success-failure
- Sampling distribution of $\hat{p} \sim Normal (\mu = p, \sigma = \sqrt{\frac{p(1-p)}{n}})$  

For computation:

| Method | Value used as $p$ |
|--------|------------|
| Confidence interval | \_\_\_\_\_\_\_\_\_\_ |
| Hypothesis test | \_\_\_\_\_\_\_\_\_\_ |


## Final Connection

Complete the sentence:

> Statistical inference for a proportion works because...


\newpage

# Study design: Sample size calculation

How can we know how large of a sample $n$ we need to collect? If we know the maximum margin of error we want, we can solve for $n$. 

Suppose we were setting up our study and wanted to know how big of a sample is required to ensure the margin of error is smaller than 0.04 using a 95% confidence level for the true proportion of orange candies?

We know the margin of error for a 95% confidence interval for a proportion is $1.96 * \sqrt{\frac{p(1 - p)}{n}} < 0.04$.
It turns out that the margin of error is largest when $p$ is 0.5, so we typically use this worst case value if no estimate of the proportion is available.

\begin{equation*}
  \begin{split}
    1.96 * \sqrt{\frac{p(1 - p)}{n}} &< 0.04\\
    1.96 * \sqrt{\frac{0.5 \times (1 - 0.5)}{n}} &< 0.04\\
    1.96 * \sqrt{\frac{0.25}{n}} &< 0.04\\
    \sqrt{\frac{0.25}{n}} &< 0.02041\\
    \frac{0.25}{n} &< 0.02041^2\\
    \frac{0.25}{n} &< 0.00042\\
    n & >  \frac{0.25}{0.00042}\\
    n & >  595.24\\
  \end{split}
\end{equation*}

So we would need a minimum of 596 candies to have a 95% confidence interval with a margin of error of 0.04.