---
title: "MATH 201: Lecture 6a Handout"
subtitle: "5.2 Confidence Intervals for a Proportion"
format: 
  pdf:
    echo: false
    message: false
    warning: false
    geometry:
      - top=0.5in
      - left=1in
      - right=1in
---

```{r}
library(tidyverse)
options(scipen = 999)

set.seed(24435)
```

\newcommand{\blank}{\rule{2.5cm}{0.15mm}}

\vspace*{-2cm}

Name:\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ Date:\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

## Learning goals for today

By the end of today, you should be able to:

- Explain why a confidence interval is more informative than a point estimate.
- Construct a 95% confidence interval for a population proportion.
- Interpret what "95% confident" means.
- Distinguish between sample size (n) and number of samples.
- Explain how sample size affects interval width.

# 1. Why Not Just Report One Number?

Suppose aviation regulators are studying aircraft safety. They want to estimate the **proportion of flights that result in a major incident**, $p$.

If they take a sample of flights and compute $\hat{p} = 0.000002$, do you just accept that value as $p$?

What if we took another sample of flights and computed $\hat{p} = 0.003$? We know we could get a different value because of variability (randomness) from sampling.

Instead of reporting one number (a point estimate), we want a range of plausible values:

$$\text{Estimate} \pm \text{Margin of Error}$$

What margin of error would you be okay with for the estimate of $p$, the number of flights that result in major incident? 

\vspace{1cm}

From the Central Limit Theorem we know if conditions are met:
$$
\hat p \sim N(\mu=p, \sigma = \sqrt{\frac{p(1 - p)}{n}})
$$
From the Empirical Rule we also know 95% of the normal distribution is within two standard deviations of the mean (technically 95.45% to be exact). So


$$
\hat{p} \pm 1.96 \times SE,
$$

where $SE = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$. This range is called a **95% confidence interval**.

\newpage

The International Air Transport Association (IATA) provides a count of commercial airline flights and incidents/accidents in 2024 (<https://www.iata.org/en/pressroom/2025-releases/2025-02-26-01/>). From the IATA 2024 Annual Safety Report: 40.6 million total flights in 2024 with 46 total accidents (crashes or serious incidents defined in their accident classification) in 2024. Let's use this reports as our sample.

Compute: 

- $n$ = 
- $\hat{p}$ = 

- $SE$ = 

- 95% confidence interval margin of error = 

- 95% confidence interval = 

Fill in the interpretation:

We are 95% confidence that the true accident rate among flights in 2024 was between ____________ and __________ accidents per million flights.

## What Does 95% Confident Mean?

If we repeated sampling forever and built a 95% confidence interval each time:

- About 95% of intervals would contain the true population proportion.
- About 5% would miss.

It does NOT mean:

"There is a 95% probability that this specific interval contains p."

The parameter is fixed. The interval is random.


## Polling example

Since we typically only have one sample, it can be hard to understand the idea of confidence intervals. Here we will simulate data to try to better understand confidence intervals.

Assume the true population proportion is: $p = 0.60$. We take a sample with $n = 100$.

```{r}
p_true <- 0.60
n <- 100

sample_data <- rbinom(n, 1, p_true)
p_hat <- mean(sample_data)
SE <- sqrt(p_hat*(1-p_hat)/n)
ME <- 1.96*SE

conf_int1 <- round(c(p_hat - ME, p_hat + ME), 2)
```

This gives us one confidence interval $0.6 \pm 1.96 \sqrt{0.6 (1 - 0.6) / n}$, aka (`r conf_int1[1]`, `r conf_int1[2]`). But what if we repeat the polling?

It is crucial to distinguish between:

- **Sample size (n)** = number of observations in a sample (e.g., people in one poll)  
- **Number of samples** = how many separate samples we collect (e.g., number of times we repeat polling)  

We now simulate 25 different polls, each with n = 100.

```{r}
num_samples <- 25
n <- 100

results <- data.frame(
  sample = 1:num_samples,
  p_hat = NA,
  lower = NA,
  upper = NA
)

for(i in 1:num_samples){
  sample_data <- rbinom(n, 1, p_true)
  p_hat <- mean(sample_data)
  SE <- sqrt(p_hat*(1-p_hat)/n)
  ME <- 1.96*SE
  
  results$p_hat[i] <- p_hat
  results$lower[i] <- p_hat - ME
  results$upper[i] <- p_hat + ME
}
```


# 4. Visualizing 25 Confidence Intervals

Each horizontal line represents one poll's confidence interval. The dashed vertical line shows the true population proportion.

```{r}
#| fig-width: 5
#| fig-height: 3
ggplot(results, aes(x = p_hat, y = sample)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_vline(xintercept = p_true, linetype = "dashed") +
  labs(
    title = "25 Different Samples (n = 100 Each)",
    x = "Sample Proportion (with 95% CI)",
    y = "Sample Number"
  ) +
  theme_minimal()
```

What do you observe?

A 95% confidence level refers to the **long-run success rate** of the method. Approximately 95% of the confidence intervals should contain the true proportion $p$.


## Effect of sample size

Now let's see what increasing $n$ does. Under the same conditions, let's now have $n = 500$:

```{r}
#| fig-width: 5
#| fig-height: 3
num_samples <- 25
n <- 500

results <- data.frame(
  sample = 1:num_samples,
  p_hat = NA,
  lower = NA,
  upper = NA
)

for(i in 1:num_samples){
  sample_data <- rbinom(n, 1, p_true)
  p_hat <- mean(sample_data)
  SE <- sqrt(p_hat*(1-p_hat)/n)
  ME <- 1.96*SE
  
  results$p_hat[i] <- p_hat
  results$lower[i] <- p_hat - ME
  results$upper[i] <- p_hat + ME
}

ggplot(results, aes(x = p_hat, y = sample)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_vline(xintercept = p_true, linetype = "dashed") +
  labs(
    title = "25 Different Samples (n = 500 Each)",
    x = "Sample Proportion (with 95% CI)",
    y = "Sample Number"
  ) +
  theme_minimal() +
  xlim(c(0.4, 0.8))
```

What changed? 
