---
title: "MATH 201: Lecture 7b Handout"
subtitle: "5.3 Hypothesis Testing for a Proportion (Continued)"
format: 
  pdf:
    echo: false
    message: false
    warning: false
    geometry:
      - top=0.5in
      - left=1in
      - right=1in
---

```{r}
library(tidyverse)
options(scipen = 999)
```

\newcommand{\blank}{\rule{2.5cm}{0.15mm}}

\vspace*{-2cm}

Name:\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ Date:\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

## Learning goals for today

By the end of this lecture, you should be able to:

- Define and interpret **Type I and Type II errors**
- Explain the meaning of the **significance level** ($\alpha$)
- Describe what a **p-value** represents in context
- Conduct a **one-proportion hypothesis test** using the p-value method
- Distinguish between **statistical significance** and **practical significance**


## Decision Errors in Hypothesis Testing

When we perform a hypothesis test, we make one of two decisions: Reject $H_0$ or Fail to reject $H_0$
But we might be wrong. There are two possible truths: $H_0$ is true or $H_A$ is true  

This creates four possible outcomes:

| Truth | Decision | Result |
|--------|----------|--------|
| $H_0$ true | Fail to reject $H_0$ | Correct decision |
| $H_0$ true | Reject $H_0$ | **Type I Error** |
| $H_A$ true | Reject $H_0$ | Correct decision |
| $H_A$ true | Fail to reject $H_0$ | **Type II Error** |


**Type I Error:** When we reject $H_0$ even though $H_0$ is actually true.

\vspace{1.25in}


**Type II Error:** When we fail to reject $H_0$ even though $H_A$ is actually true.

\vspace{1.25in}

Example: Courtroom Analogy

- $H_0$: Defendant is innocent  
- $H_A$: Defendant is guilty  

Type I Error:

\vspace{0.75in}

Type II Error: 

\vspace{0.75in}

Which error does the legal system try harder to avoid? Why?

\vspace{0.75in}


## Significance Level ($\alpha$)

The **significance level**, written $\alpha$, is the probability of making a Type I Error.

If $\alpha = 0.05$, this means:

If the null hypothesis is true, we will incorrectly reject it about _______% of the time in the long run.

Circle one: Smaller $\alpha$ → *more* / *less* cautious about rejecting $H_0$

Reflection: If rejecting $H_0$ incorrectly would cause serious harm, should $\alpha$ be small or large?

\vspace{1cm}


## From Confidence Intervals to p-values

Previously, we used confidence intervals to determine whether a null value was plausible. Today, we use p-values.
The **p-value** is the probability of observing data **at least as extreme as** what we observed, assuming that null hypothesis is true.

- Small p-value → observed data are unlikely if $H_0$ is true → strong evidence against $H_0$  
- Moderate to large p-value → observed data are plausible if $H_0$ is true → no evidence against $H_0$


### Example: Testing a Majority

Kim et al. (2004) studied the measles-rubella vaccination-rates in Korea, comparing the proportion of children susceptible to measles with the World Health Organization target proportion (for children aged 5 to 9 years old: 10%). The aim was to test if the proportion of Korean children susceptible to measles in the population was 
10% or lower (i.e., better). In the study, 55 children out of 972 were susceptible to measles.


#### Step 1: State Hypotheses

If we want to test whether there is a majority either way:

\begin{equation*}
  \begin{split}
    H_0& :\\
    H_A& :
  \end{split}
\end{equation*}


#### Step 2: Check Conditions

Independence: 

\vspace{1in}

Success–failure condition (use $p_0$ for $p$):

\vspace{1in}

#### Step 3: Calculate evidence

$n$ =  

$\hat{p}$ = 

$SE = \sqrt{\frac{p_0(1 - p_0)}{n}} =$

$z = \frac{\hat{p} - p_0}{SE} =$

This means the sample proportion is ______ standard errors below / above the null value ($p_0$).


Compute the p-value (always make a sketch).

- p-value:  

- Probability of observing data as or more extreme, if $H_0$ is true:

#### Step 4: Decide & Conclude

Decision (i.e. Is our observed data unlikely if the $H_0$ is true?): 

\vspace{1.5in}

Conclusion in context:

\vspace{1.5in}

## Statistical vs Practical Significance

With very large sample sizes, even tiny differences from the null value can produce very small p-values. This is called **statistical significance** (i.e. p-value < significance level).

But we must also ask: Is the difference large enough to matter in the real world? This is called **practical significance** (e.g. $\hat{p}$ vs $p_0$).

- For example: Imagine for the vaccination data we had a sample size $n = 10,000$ and $\hat{p} = 0.1005$. We may find evidence against $H_0$ (statistical significance), but the difference between 10.05% ($\hat{p}$) and 10% ($p_0$) is tiny, indicating low practical significance.
- Always step back and think about the practical significance of the results!

## The 3 types of alternative hypotheses $H_A$

\vspace{0.5in}

```{r}
#| fig-align: center
#| fig-height: 1
#| fig-width: 7
# Create x values
x <- seq(-4, 4, length.out = 500)

# Standard normal density
df <- data.frame(
  x = x,
  y = dnorm(x)
)

blank_normal <- ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.3) +

  # Only one tick at 0 labeled mu
  # scale_x_continuous(
  #   breaks = 0,
  #   labels = "",
  #   expand = c(0.02, 0.02)
  # ) +
  
  # Remove y-axis ticks and labels
  scale_y_continuous(breaks = NULL) +
  
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  )

gridExtra::grid.arrange(
  blank_normal,
  blank_normal,
  blank_normal,
  nrow = 1
)
```

\vspace{1in}

## Example: Home-field advantage

In the 2019/2020 English Premier League (EPL), the home team won 91 games, the away team won 67 games, and another 50 games were draws. Determine if there is evidence that the home team wins more often than 50% (i.e., that there is a home-field advantage), using a hypothesis test with a significance level of 0.1.

#### Step 1: State Hypotheses

If we want to test whether there is a majority either way:

\begin{equation*}
  \begin{split}
    H_0& :\\
    H_A& :
  \end{split}
\end{equation*}


#### Step 2: Check Conditions

Independence: 

\vspace{1in}

Success–failure condition (use $p_0$ for $p$):

\vspace{1in}

#### Step 3: Calculate evidence

$n$ =  

$\hat{p}$ = 

$SE = \sqrt{\frac{p_0(1 - p_0)}{n}} =$

$z = \frac{\hat{p} - p_0}{SE} =$

This means the sample proportion is ______ standard errors below / above the null value ($p_0$).


Compute the p-value (always make a sketch).

- p-value:  

- Probability of observing data as or more extreme, if $H_0$ is true:

#### Step 4: Decide & Conclude

Decision (i.e. Is our observed data unlikely if the $H_0$ is true?): 

\vspace{3in}

Conclusion in context:

\vspace{3in}


Does there appear to be practical significance (aka a practical difference)?

<!-- Examples pulled from: https://peterkdunn.github.io/SRM-Textbook/TestOneProportion.html -->
